\input{../written/header}
\usepackage{outline}

\begin{document}
\begin{outline}
	\item \textbf{Personal background intro?}  % informal
	\begin{outline}
		\item Major / CS background - interests?
		\item Assistant Research Engineer 2017-2018
	\end{outline}
	% \item \textbf{Abstract}
	\item \textbf{Background}
	\begin{outline}
		\item Importance of sequence alignment.
		\item Genomic data contains artifacts (frameshifts, early stop codons).
		\item Current aligners are note equipped to handle this issue.
		\item Within codon indels figure.
	\end{outline}
	\item \textbf{Aims} - short intro: to solve this issue I propose to develop
				COATi (etc).
	\item \textbf{Aim 1}
	\begin{outline}
		\item Quick HMM recap.
		\item FSTs and composition.
		\item Probability figure.
		\item Evolution FST.
		\item Substitution model.
		\item DP? (Brief).
	\end{outline}
	\item \textbf{Aim 2}
	\begin{outline}
		\item Artifacts?
		\item High quality and low quality (heterogeneous) approach.
		\item Marginalized substitution model.
		\item Ambiguous data.
		\item Complex frameshift modeling.
		\item Biological frameshifts?
	\end{outline}
	\item \textbf{Aim 3}
	\begin{outline}
		\item Importance of estimating parameter values.
		\item EM.
		\item Substitution and indel model parameters.
		\item Validation?
	\end{outline}
	\item \textbf{Preliminary Results}
	\begin{outline}
		\item Quick summary of what's done.
		\item Software comparison table and analysis description.
		\item Shortcomings of aligners?
		\item Memory \& runtime plots.
	\end{outline}
	\item \textbf{Future work}
	\item \textbf{Expected graduation date?}
	\item References
\end{outline}
% \printbibliography
\end{document}
